Isomorphic Interpreters from Logically Reversible Abstract Machines

### 2012-03-16 iso-int/rc-iso-int.tex
last section and withim page limit

Let us apply these steps to the slightly more interesting example of an
adder:
\begin{multicols}{2}  
%subcode{bnf} include main
% Numbers, n, m,p = 0 | n + 1
% Machine states = <e, e, e>

%subcode{bnf} include main
% Start state = <n, n, 0>
% Stop State = <0, n, n>
\end{multicols}
%subcode{opsem} include main
% <n+1, p, m> |--> <n, p+1, m+1>

@@

\begin{center}
\scalebox{0.9}{
  \includegraphics{diagrams/adder3.pdf}
}
\end{center}

@@

\begin{center}
\scalebox{0.9}{
  \includegraphics{diagrams/adder4.pdf}
}
\end{center}

@@

\begin{multicols}{2}
%subcode{bnf} include main
% Tree, t = L n | N t t
% Tree Contexts, c = Lft c t | Rgt t c | []
%
% Machine states = <t, c> | {[c, t]}
% Start state = <t, []>
% Stop State = {[[], t]}

%subcode{opsem} include main
% <L n, c> &|-->& {[c, L (n+1)]}
% <N t1 t2, c> &|-->& <t1, Lft c t2>
% {[Lft c t2, t1]} &|-->& <t2, Rgt t1 c>
% {[Rgt t1 c, t2]} &|-->& {[N t1 t2, c]}

\end{multicols}

@@

\begin{multicols}{2}
%subcode{bnf} include main
% Machine states = <bool, t, c>
%
% Start state = <T, t, []>
% Stop state =  <F, t, []>

%subcode{opsem} include main
%! columnStyle = rclr
% <T, L n, c> &|-->& <F, L (n+1), c> &~~~~~ rule 1
% <T, N t1 t2, c> &|-->& <T, t1, Lft c t2> & rule 2 
% <F, t1, Lft c t2> &|-->& <T, t2, Rgt t1 c> & rule 3
% <F, t2, Rgt t1 c> &|-->& <F, c, N t1 t2> & rule 4
\end{multicols}

@@

@@

>>>>

Let us apply these steps to the slightly more interesting example of an
adder:

\vspace{-20pt}
\begin{small}
\begin{multicols}{2}  
%subcode{bnf} include main
% Numbers, n, m,p = 0 | n + 1
% Machine states = <e, e, e>

%subcode{bnf} include main
% Start state = <n, n, 0>
% Stop State = <0, n, n>
\end{multicols}
\vspace{-15pt}
%subcode{opsem} include main
% <n+1, p, m> |--> <n, p+1, m+1>
\end{small}

@@

\begin{center}
\scalebox{0.7}{
  \includegraphics{diagrams/adder3.pdf}
}
\end{center}

@@

\begin{center}
\scalebox{0.7}{
  \includegraphics{diagrams/adder4.pdf}
}
\end{center}

@@

\vspace{-15pt}
\begin{small}
\begin{multicols}{2}
%subcode{bnf} include main
% Tree, t = L n | N t t
% Tree Contexts, c = Lft c t | Rgt t c | []
%
% Machine states = <t, c> | {[c, t]}
% Start state = <t, []>
% Stop State = {[[], t]}

%subcode{opsem} include main
% <L n, c> &|-->& {[c, L (n+1)]}
% <N t1 t2, c> &|-->& <t1, Lft c t2>
% {[Lft c t2, t1]} &|-->& <t2, Rgt t1 c>
% {[Rgt t1 c, t2]} &|-->& {[N t1 t2, c]}
\end{multicols}
\end{small}

@@

\begin{small}
\begin{multicols}{2}
%subcode{bnf} include main
% Machine states = <bool, t, c>
%
% Start state = <T, t, []>
% Stop state =  <F, t, []>

%subcode{opsem} include main
%! columnStyle = rclr
% <T, L n, c> &|-->& <F, L (n+1), c> &~~~~~ rule 1
% <T, N t1 t2, c> &|-->& <T, t1, Lft c t2> & rule 2 
% <F, t1, Lft c t2> &|-->& <T, t2, Rgt t1 c> & rule 3
% <F, t2, Rgt t1 c> &|-->& <F, c, N t1 t2> & rule 4
\end{multicols}
\end{small}

@@

In the final construction we present a metacircular interpreter for
{{langRevT}} written in {{langRevT}}. This is a a non-trivial abstract
machine with several cases, but the deriving the corresponding
{{langRevT}} interpreter follows in exactly the same way as
before. Here is a logically reversible small-step abstract machine for
{{langRevT}}.

\begin{small}

@@

\end{small}

The isomorphic interpreter thus derived is shown below, just before
the `flipping' step (which makes it harder to read). Each wire has
been annotated with the reduction rule it corresponds to making it
easier to compare with the abstract machine.

\noindent
A couple of important points:

\begin{enumerate}

\item The definition here shows only the composition combinators;
  which are the intersting cases. All the primitive isomorphisms are
  hidden in the `{{iso(v)}}' application in {{rule 1}}. This should be
  read as `tranformation {{v}} according to the isomorphism
  {{iso}}'. Implementing this is straightformard and is hence skipped.

\item The abstract machine definition is not very strongly typed in
  teh following sense: we know that every {{v:b1*b2}} must have the
  form {{(v1,v2)}} however is cases like {{rule 8}} if we allow any
  {{v}} to appear the machine will be in an undefined state. The usual
  way this is handled in any interpreter is to either test to ensure
  that the runtime value has teh required form or to user s stronger
  typing discipline such as GADTs.We abstract away from this choice in
  this presentation.
\end{enumerate}
